Discovering causal factors with back-and-forth regression

### 2019-05-18 paper/paper.tex

In this paper, we introduce double regression, which combines direct and inverse approaches.
%
We first predict X from Y, as in the inverse approach, and retrieve a regularised least square estimate $\hat X$.
%
Then, we calculate a regularised least square estimator of $\hat X$ from X.
%
If the active features of X are represented as EX, with E a binary diagonal, this provides a good approximation of E, robust to noise and environment change.
%
The main purpose of double regression is feature extraction, but it can be turned into a regression method by filtering X on the selected features.

In the next sections, we describe double regression, prove that it asymptotically extracts E, both in the linear and non linear case, and that it can be understood as a special case of canonical component analysis (cca).
%
By experimenting with simulated data, we demonstrate its robustness and use cases, and show that it compares favourably with state of the art techniques for regression and feature extraction.
%
Finally, we illustrate its use on an example from neuroscience.

\section{Double regression}

@@

Double regression consists in two steps.
%
First, we perform a least squares estimation of X from Y, that is we find $\hat G$ such that $\hat X=\hat G Y$ is the least square predictor of X.
%
This recovers an approximation of the causes of Y in X.
%
Then, we perform a least squares estimation of $\hat X$ from X, and determine $\hat H$ such that $\hat {\hat X}=\hat H X$ is the least square predictor of $\hat X$.
%
In matrix form, we have $\hat G=(Y'Y)^{-1} Y'X$, and
\begin{equation} \hat H=(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X\end{equation}
Under good conditions (close to linear dependence between X and Y, and covariance matrices of X and Y well conditioned), this recovers a close approximation to E.

In practice, since the covariance matrices of Y and X might exhibit multicollinearity, we use regularisation in both matrix inversions.
%
This is done by using ridge regression, and determining at each step the regularisation parameter by cross validation.
%
In the above formula, this amounts to replacing $(X'X)^{-1}$ and $(Y'Y)^{-1}$ by $(X'X+\lambda)^{-1}$ and $(Y'Y+\mu)^{-1}$.
%
Also, to prevent spurious correlations between the two regressions (should they be performed on the same sample), our original sample is randomly split into two halves, and each step is performed over one of them, effectively decorrelating the two steps.
%
This is repeated over several (10) splits, and the resulting $\hat H$ are averaged.
%
This bagging compensates for the reduction in sample size caused by splitting.

@@

To select large elements of $\hat H$, the traditional approach in regression analysis consists in testing whether those coefficient differ from zero (or are larger from zero since we are only interested in large positive coefficients).
%
This is usually done via Student t-test of variants of it.
%
This will not work here, as double regression introduces, through two regularisations, a significant amount of positive bias on the diagonal, which depends in a complex manner on noise, dimensionality and conditioning of X and Y.
%
Instead, we treat feature extraction as a one dimensional binary clustering problem: classify n positive real values as "large" and "small" (under the hypothesis that they are not all small or all large).

@@

This extract from $\hat H$ a binary diagonal $\hat E$.
%
To use double regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $\hat E X$.

Summarizing, double regression is performed as follows:
\begin{enumerate}
\item Randomly split the data sample.
\item From the first half sample, using crossalidated ridge regression, find $\hat G$, the regularised least square regression of X from Y: $\hat G=(Y'Y)^{-1} Y'X$.
\item From the second half sample, calculate $\hat X = \hat G Y$, and find $\hat H$, the regularized least square regression of $\hat X$ from X: $\hat H=(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X$.
\item Repeat steps 1 to 3 and average the results as $\hat H$.
\item Using the Sonquist and Morgan criterion, select the large diagonal elements of $\hat H$, set them to one and the rest of $\hat  H$ to zero.
\item Use the resulting matrix $\hat E$ as an estimator of E, the active features in the causal process
\item Alternatively, use $\hat E$ to predict Y from X, by performing a regularised least square regression of Y from $\hat E X$.
%

@@

\subsection{Asymptotic behaviour - linear case}
As discussed in the introduction, our problem can be formulated (up to a redifinition of the features of X) as $Y=f(EX+N)$, with F an unknown function.
%
If F is linear we can rewrite it as $Y = F(EX + N)$, with X an (n, dx) matrix of possible causes, E a (dx,dx) binary diagonal matrix selecting active features of X.
%
N a (dx,n) homoscedastic noise matrix, F an unknown (possibly full rank) (dy,dx) transformation corresponding to the measuring apparatus, and Y a (n,dy) matrix of measured effects.

Since N is centred, the first least square regression (X from Y), amounts to finding
\begin{equation}
\begin{aligned}
\min_G \left \| X-GY \right \|^2 &= \min_G \left \| X - GF(EX+N)\right\|^2 \\
&{}= \min_G \left \| (I-GFE)X\right\| ^2 + \left \| GFN\right \| ^2\\
&{}= \min_G \alpha \left \| I-GFE\right\| ^2  + \beta \left \| G\right \| ^2\\
\end{aligned}
\end{equation}
with $\alpha$ and $\beta$ positive constants, the latter equal to the norm of FN. We are balancing two minimisation, with $/alpha$ close to zero, the solution is the null matrix, with $\beta$ close to zero, it is the pseudo inverse of $FE$.  
%

@@

The second regression is from X to the predicted value of X in the above, or $\hat G Y$
\begin{equation}
\begin{aligned}
arg \min_H \left \| k(FE)^{\dagger}Y - HX \right \|^2 &= arg \min_H \left \| k(FE)^{\dagger}F(EX+N) - HX \right \|^2 \\
&= arg \min_H \left \| (k(FE)^{\dagger}(FE)-H)X \right \| ^2 + \left \| k(FE)^{\dagger}FN \right \| ^2\\
&= arg \min_H \left \| (k(FE)^{\dagger}(FE)-H)X \right \| ^2\\
&=k(FE)^{\dagger}(FE)
\end{aligned}
\end{equation}
Thus, in the linear case, double regression retrieves $k(FE)^{\dagger}(FE)$. In many cases, if v s the variance of $FN$, k can be approximated as $k= {1\over{1+v}}$, it quantifies the amount of noise added to the causal process, and the difficulty of recovering the causal features.

@@

\section{Related works}
\subsection{Double regression as a special case of canonical component analysis}
Formula (1) shows that double regression can be seen as a special case of canonical component analysis (CCA).
%
In CCA, we are given two matrixes X and Y (or size N,p and N,q) that describe different features measured on the same data sample, and characterise their proximity (that is, between the subspaces spanned by their columns) as the correlation between linear combinations of X and Y.

@@

If (1) is diagonal, for instance in the model presented above, the features of X are eigenvectors.
%
If all eigenvalues are either 0 or 1, since they correspond to the squared cosine of the angle between the corresponding feature and the subspace spanned by Y, all the features are either part of the Y subspace (eigenvalues of one) or orthogonal to it (for zero eigenvalues).
%
Thus, double regression amounts to CCA under the constraint that the subspaces spanned by X and Y are perpendicular, X is spanned by vectors of Y and vectors orthogonal to Y, and that the parallel and orthogonal components of X are spanned by disjoint sets of features of X.
%
The closer we are to this hypothesis, the closer the results of CCA and double regression will be.

There are differences between CCA and double regression.
%
First, since our objective is not to characterise the proximity between X and Y, but just to extract causal features, we can dispense with the last step of CCA (diagonalisation of (1)).
%
Second, since we operate in a noisy environment, we use regularisation, introducing bias in the correlation calculations.
%
Finally, whereas CCA wants to leverage all correlations between X and Y, we filter some of them, through splitting and bagging.
%
In other words, whereas we are working from the same formulae as CCA, we use them in different ways, for different purposes

@@

\section{Experiments - simulated data}
Experiments on simulated data serve three purposes: compare double regression with other known techniques (CCA, OLS, PLS...) both as a regression and a feature extraction method, measure its performance for various values of its parameters and provide evidence for our claims about stability, invariance to environment, and behavior as a causal detector.

@@

We evaluate the performance of double regression in two different roles : as a feature extractor recovering E, and as a regression method, predicting Y from X over a test sample.
%
For feature extraction, the quality of fit is the number of disagreements between our estimator and the diagonal of E (ie, their Hamming distance, which should be minimised).
%
For regression, we use a test set of N additional measurements, and calculate the R-squared ratio of variances (which should be maximised).
%
Performance indicators are calculated by averaging the results of several experiments.

\subsection{Feature extraction - linear case}
All other parameters being equal, the quality of feature extraction, measured as the discrepancy between features extracted by double regression ($\hat E$) and the diagonal of E, improves with signal to noise ratio.
%
The decrease is usually very steep until a Hamming distance under 2 is achieved.
%
Figure 1 presents a number of such curves for varying values of dx=dy=dz, nc=30 and N=1000.

@@

For feature extraction, double regression either works very well (Hamming distance less than 2) or very badly, with a sharp change over small variations of the parameters.
%
The following graphs show the boundary between values of dx and dz that allow for features extractions (dx lower than boundary, or dz larger), for different values of dy (level curves on the graph), and different values of signal to noise ratio and number of active variables.

@@

For values of ...
%
in ..., an empirical formula can be provided.
%
Double regression allows for good feature extraction if :

@@

\subsection{Regression }
For regression, we want to understand how feature extraction helps predict Y from X.
%
To this effect, we compare four approaches to regression:
\begin{itemize}
\item standard ridge regression of Y from X, where no features are extracted this is our baseline
\item weighted ridge regression, of Y from $\hat H X$, where features of X are weighted according to the diagonal of $\hat H$
\item filtered ridge regression, of Y from $\hat E X$, where active features are first extracted (as above), and Y is predicted from the extracted features only
\item perfect filtered (oracle) ridge regression of Y from EX, where E is supposed known
\end{itemize}
Over a large set of parameters, we observe that filtered regression is on average better than weighted, which is on average better than standard regression.
%
This demonstrates that double regression can be used as a regression method, over the areas in parameter space where feature extraction performs well.

@@

The rationale behind such tests is that whereas double regression, being a feature extraction method, might not be as efficient as dedicated prediction tools, its focus on causal elements (and the removal of non causal features before carrying out regression) might cause it to be more resistant to changes in environments.

>>>>

In this paper, we introduce back and forth regression, which combines direct and inverse approaches.
%
We first predict X from Y, as in the inverse approach, and retrieve a regularised least square estimate $\hat X$.
%
Then, we calculate a regularised least square estimator of $\hat X$ from X.
%
If the active features of X are represented as EX, with E a binary diagonal, this provides a good approximation of E, robust to noise and environment change.
%
The main purpose of back and forth regression is feature extraction, but it can be turned into a regression method by filtering X on the selected features.

In the next sections, we describe back and forth regression, prove that it asymptotically extracts E, both in the linear and non linear case, and that it can be understood as a special case of canonical component analysis (cca).
%
By experimenting with simulated data, we demonstrate its robustness and use cases, and show that it compares favourably with state of the art techniques for regression and feature extraction.
%
Finally, we illustrate its use on an example from neuroscience.

\section{Back an forth regression}

@@

Back and forth regression consists in two steps.
%
First, we perform a least squares estimation of X from Y, that is we find $\hat G$ such that $\hat X=\hat G Y$ is the least square predictor of X.
%
This recovers an approximation of the causes of Y in X.
%
Then, we perform a least squares estimation of $\hat X$ from X, and determine $\hat H$ such that $\hat {\hat X}=\hat H X$ is the least square predictor of $\hat X$.
%
In matrix form, we have $\hat G=(Y'Y)^{-1} Y'X$, and
\begin{equation} \hat H=(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X\end{equation}
Under good conditions (close to linear dependence between X and Y, and covariance matrices of X and Y well conditioned), this recovers a scaled approximation to E.

In practice, since the covariance matrices of Y and X might exhibit multicollinearity, we use regularisation in both matrix inversions. This amounts to replacing $(X'X)^{-1}$ and $(Y'Y)^{-1}$ by $(X'X+\lambda)^{-1}$ and $(Y'Y+\mu)^{-1}$, and is done by using ridge regression, regularisation parameters being estimated by leave one out cross validation.
%
Also, to prevent spurious correlations between the two steps (should they be performed on the same sample), our original sample is randomly split into two halves, and each step is performed over one of them, effectively decorrelating the two steps.
%
This is repeated over several (10) splits, and the resulting $\hat H$ are averaged.
%
This bagging compensates for the reduction in sample size caused by splitting.

@@

To select large elements of $\hat H$, the traditional approach in regression analysis consists in testing whether its coefficients differ from zero. This is usually done via variants of Student t-test.
%
This will not work here, as back and forth regression introduces, through two regularisations, a significant amount of positive bias on the diagonal, which depends in a complex manner on noise, dimensionality and conditioning of X and Y.
%
Instead, we treat feature extraction as a one dimensional binary clustering problem: classify n positive real values as "large" and "small" (under the hypothesis that they are not all small or all large).

@@

This extract from $\hat H$ a binary diagonal $\hat E$.
%
To use back and forth regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $\hat E X$.

Summarizing, back and forth regression is performed as follows:
\begin{enumerate}
\item Randomly split the data sample.
\item From the first half sample, using crossalidated ridge regression, find $\hat G$, the regularised least square regression of X from Y: $\hat G=(Y'Y)^{-1} Y'X$.
\item From the second half sample, calculate $\hat X = \hat G Y$, and find $\hat H$, the regularized least square regression of $\hat X$ from X: $\hat H=(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X$.
\item Repeat steps 1 to 3 and average the results as $\hat H$.
\item Using the Sonquist and Morgan criterion, select the large diagonal elements of $\hat H$, set them to one and the rest of $\hat  H$ to zero.
\item Use the resulting matrix $\hat E$ as an estimator of E, the active features in the causal process
\item Alternatively, use $\hat E$ to predict Y from X, by performing a regularised least square regression of Y from $\hat E X$.
%

@@

\subsection{Asymptotic behaviour - linear case}
As discussed in the introduction, our problem can be formulated (up to a redefinition of the features of X) as $Y=f(EX+N)$, with F an unknown function.
%
If F is linear we can rewrite it as $Y = F(EX + N)$, with X an (n, dx) matrix of possible causes, E a (dx,dx) binary diagonal matrix selecting active features of X.
%
N a (dx,n) homoscedastic noise matrix, F an unknown (possibly full rank) (dy,dx) transformation corresponding to the measuring apparatus, and Y a (n,dy) matrix of measured effects.

The two steps of back and forth regression amounts to finding $\hat G=\arg \min_G \left \| X-GY \right \|^2$ and $\hat H =\arg \min_H \left \| \hat GY - HX \right \|^2$ (all matrix norms henceforth being Frobenius norms).

The second step is straightforward, replacing Y and since N is centred, we have
\begin{equation}
\begin{aligned}
\arg \min_H \left \| \hat GY - HX \right \|^2 &=\arg  \min_H \left \| \hat GF(EX+N) - HX \right \|^2 \\
&=\arg \min_H \left \| (\hat G(FE)-H)X \right \| ^2 + \left \| \hat GFN \right \| ^2\\ 
&= \arg \min_H \left \| (\hat G(FE)-H)X \right \| ^2\\
&=\hat G (FE)
\end{aligned}
\end{equation}

For the first step, we have 
\begin{equation}
\begin{aligned}
\min_G \left \| X-GY \right \|^2 &= \min_G \left \| X - GF(EX+N)\right\|^2 \\
&= \min_G \left \| (I-GFE)X\right\| ^2 + \left \| GFN\right \| ^2\\
&= \min_G \sum_{i} \left \| (I-GFE)X_{i}\right\| ^2 + \left \| GFN_{i}\right \| ^2\\
&{}= \min_G \alpha \left \| I-GFE\right\| ^2  + \beta \left \| G\right \| ^2\\
\end{aligned}
\end{equation}
with $\alpha$ and $\beta$ positive constants, the latter equal to the norm of FN. We are balancing two minimisation, with $\alpha$ close to zero, the solution is the null matrix, with $\beta$ close to zero, it is the pseudo inverse of $FE$.  
%
FC: lost here . I'd like to prove that we recover a scales version $(FE)^\dagger D$ with D a diagonal scaling over dimensions of X, according to signal noise ratio, covariance of X and covariance of N. This is what we sort of get from

SVD of X, USV, so the minimum becomes

@@

Thus, in the linear case, back and forth regression retrieves $k(FE)^{\dagger}(FE)$. In many cases, if v s the variance of $FN$, k can be approximated as $k= {1\over{1+v}}$, it quantifies the amount of noise added to the causal process, and the difficulty of recovering the causal features.

@@

\section{Related works}
\subsection{back and forth regression as a special case of canonical component analysis}
Formula (1) shows that back and forth regression can be seen as a special case of canonical component analysis (CCA).
%
In CCA, we are given two matrixes X and Y (or size N,p and N,q) that describe different features measured on the same data sample, and characterise their proximity (that is, between the subspaces spanned by their columns) as the correlation between linear combinations of X and Y.

@@

If (1) is diagonal, for instance in the model presented above, the features of X are eigenvectors.
%
If all eigenvalues are either 0 or 1, since they correspond to the squared cosine of the angle between the corresponding feature and the subspace spanned by Y, all the features are either part of the Y subspace (eigenvalues of one) or orthogonal to it (for zero eigenvalues).
%
Thus, back and forth regression amounts to CCA under the constraint that the subspaces spanned by X and Y are perpendicular, X is spanned by vectors of Y and vectors orthogonal to Y, and that the parallel and orthogonal components of X are spanned by disjoint sets of features of X.
%
The closer we are to this hypothesis, the closer the results of CCA and back and forth regression will be.

There are differences between CCA and back and forth regression.
%
First, since our objective is not to characterise the proximity between X and Y, but just to extract causal features, we can dispense with the last step of CCA (diagonalisation of (1)).
%
Second, since we operate in a noisy environment, we use regularisation, introducing bias in the correlation calculations.
%
Finally, whereas CCA wants to leverage all correlations between X and Y, we filter some of them, through splitting and bagging.
%
In other words, whereas we are working from the same formulae as CCA, we use them in different ways, for different purposes

@@

\section{Experiments - simulated data}
Experiments on simulated data serve three purposes: compare back and forth regression with other known techniques (CCA, OLS, PLS...) both as a regression and a feature extraction method, measure its performance for various values of its parameters and provide evidence for our claims about stability, invariance to environment, and behavior as a causal detector.

@@

We evaluate the performance of back and forth regression in two different roles : as a feature extractor recovering E, and as a regression method, predicting Y from X over a test sample.
%
For feature extraction, the quality of fit is the number of disagreements between our estimator and the diagonal of E (ie, their Hamming distance, which should be minimised).
%
For regression, we use a test set of N additional measurements, and calculate the R-squared ratio of variances (which should be maximised).
%
Performance indicators are calculated by averaging the results of several experiments.

\subsection{Feature extraction - linear case}
All other parameters being equal, the quality of feature extraction, measured as the discrepancy between features extracted by back and forth regression ($\hat E$) and the diagonal of E, improves with signal to noise ratio.
%
The decrease is usually very steep until a Hamming distance under 2 is achieved.
%
Figure 1 presents a number of such curves for varying values of dx=dy=dz, nc=30 and N=1000.

@@

For feature extraction, back and forth regression either works very well (Hamming distance less than 2) or very badly, with a sharp change over small variations of the parameters.
%
The following graphs show the boundary between values of dx and dz that allow for features extractions (dx lower than boundary, or dz larger), for different values of dy (level curves on the graph), and different values of signal to noise ratio and number of active variables.

@@

For values of ...
%
in ..., an empirical formula can be provided.
%
back and forth regression allows for good feature extraction if :

@@

\subsection{Regression }
For regression, we want to understand how feature extraction helps predict Y from X.
%
To this effect, we compare four approaches to regression:
\begin{itemize}
\item standard ridge regression of Y from X, where no features are extracted this is our baseline
\item weighted ridge regression, of Y from $\hat H X$, where features of X are weighted according to the diagonal of $\hat H$
\item filtered ridge regression, of Y from $\hat E X$, where active features are first extracted (as above), and Y is predicted from the extracted features only
\item perfect filtered (oracle) ridge regression of Y from EX, where E is supposed known
\end{itemize}
Over a large set of parameters, we observe that filtered regression is on average better than weighted, which is on average better than standard regression.
%
This demonstrates that back and forth regression can be used as a regression method, over the areas in parameter space where feature extraction performs well.

@@

The rationale behind such tests is that whereas back and forth regression, being a feature extraction method, might not be as efficient as dedicated prediction tools, its focus on causal elements (and the removal of non causal features before carrying out regression) might cause it to be more resistant to changes in environments.
